---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mpitbR: A Toolbox for Calculating Multidimensional Poverty Indices in R

<!-- badges: start -->
<!-- badges: end -->

Here it is provided a package for estimating multidimensional poverty measures based on the Alkire-Foster method which mirrors the estimation procedures of the original mpitb Stata package. 


## Installation

You can install the development version of mpitbR from [GitHub](https://github.com/girelaignacio/mpitbR) or installing it from CRAN. 

```{r intallation}
# Define the package name
package_name <- "mpitbR"

# If the package is not available on CRAN, try installing it from GitHub
if (!require(package_name, character.only = TRUE)) {
  devtools::install_github("girelaignacio/mpitbR")
}

# Try installing the package from CRAN
if (!require(package_name, character.only = TRUE)) {
  install.packages(package_name)
}

```
This process may take time the first time. 

---
