---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mpitbR: A Toolbox for Calculating Multidimensional Poverty Indices in R

<!-- badges: start -->
<!-- badges: end -->

Here it is provided a package for estimating multidimensional poverty measures based on the Alkire-Foster method which mirrors the estimation procedures of the original mpitb Stata package. 


## Installation

You can install the development version of mpitbR from [GitHub](https://github.com/girelaignacio/mpitbR) or installing it from CRAN. 

```{r intallation}
# Define the package name
package_name <- "mpitbR"

# If the package is not available on CRAN, try installing it from GitHub
if (!require(package_name, character.only = TRUE)) {
  devtools::install_github("girelaignacio/mpitbR")
}

# Try installing the package from CRAN
if (!require(package_name, character.only = TRUE)) {
  install.packages(package_name)
}

```
This process may take time the first time. 


## Some comparisons with the original Stata package

Under this setting in the <tt>mpitb</tt> Stata package:

> mpitb est, name(trial01) measures(all) ///
indmeasures(all) aux(all)  ///
klist(20 33) weight(equal) ///
lframe(myresults, replace) ///
svy over(area region) ///
cotmeasures(all) cotframe(mycot, replace) tvar(t) cotyear(year)
	
Here below, we compare the results from the Stata package and this version in R. We merged both data frames with the results (point estimate, standard errors, and confidence intervals). We calculate the L1 distance between the estimates ($|\theta_{R}-\theta_{Stata}|$) by measure in the case of the "lframe" results and by ctype and annualized measure in the case of the "cotframe" results.
```{r read_the_dataframes, echo=FALSE}

# lframe testing ----------------------------------------------------------

R <- readRDS("~/lframeR.rds")
R <- R[,-which(colnames(R) == "ctype")]

Stata <- as.data.frame(haven::read_dta("~/lframe.dta"))

Stata <- Stata[, !(names(Stata) %in% c("spec","wgts","pval","tval"))]
Stata$subg[which(is.na(Stata$subg))] <- "nat"
Stata <- Stata[-( which(Stata$measure == "popsh")),]
Stata$indicator[which(Stata$indicator== "")] <- NA
Stata <- Stata[,-which(colnames(Stata) == "ctype")]
Stata <- Stata[names(R)]

R<-R[order(R$subg, R$t, R$loa,R$measure,R$k,R$indicator),]
Stata<-Stata[order(Stata$subg, Stata$t, Stata$loa,Stata$measure,Stata$k,Stata$indicator),]

merged <- merge(R,Stata,by= c("subg","t","loa","measure","k","indicator"))



# cotframe testing --------------------------------------------------------


cotR <- readRDS("~/cotframeR.rds")
cotR <- cotR[,-(which(colnames(cotR) == "yt0")) ]
cotR <- cotR[,-(which(colnames(cotR) == "yt1")) ]

cotStata <- as.data.frame(haven::read_dta("~/cotframe.dta"))
cotStata <- cotStata[, !(names(cotStata) %in% c("spec","wgts","pval","tval","yt0","yt1","t0","t1"))]
cotStata$subg[which(is.na(cotStata$subg))] <- "nat"
cotStata$indicator[which(cotStata$indicator== "")] <- NA
cotStata$ctype <- haven::zap_labels(cotStata$ctype)
cotStata$ctype[which(cotStata$ctype == 1)] <- "abs"
cotStata$ctype[which(cotStata$ctype == 2)] <- "rel"


cotStata <- cotStata[names(cotR)]

cotR<-cotR[order(cotR$subg, cotR$loa,cotR$measure,cotR$k,cotR$indicator,cotR$ctype,cotR$ann),]
cotStata<-cotStata[order(cotStata$subg, cotStata$loa,cotStata$measure,cotStata$k,cotStata$indicator, cotStata$ctype, cotStata$ann),]


cot.merged <- merge(cotR,cotStata,by= c("subg","loa","measure","k","indicator","ctype","ann"))

```

The following output shows the summary of the distribution of this comparisons (suffix ".x" refers to the estimates of the R package and ".y" to the Stata version).

```{r summaries, echo=T}
# "lframe" COMPARISONS
aggregate(abs(b.x - b.y)~ measure, merged, summary)
aggregate(abs(se.x - se.y)~ measure, merged, summary)
aggregate(abs(ll.x - ll.y)~ measure, merged, summary)
aggregate(abs(ul.x - ul.y)~ measure, merged, summary)
# "cotframe" COMPARISONS
aggregate(abs(b.x - b.y)~ ctype+ann, cot.merged, summary)
aggregate(abs(se.x - se.y)~ ctype+ann, cot.merged, summary)
aggregate(abs(ll.x - ll.y)~ ctype+ann, cot.merged, summary)
aggregate(abs(ul.x - ul.y)~ ctype+ann, cot.merged, summary)

```

All the results are very close with some exceptions found in few estimates of the standard errors of the relative changes over time measures. Since the point estimates ("b") are close and Delta method is used for calculating the standard errors, here there may be some differences between the "survey" library in R and Stata
	
---
