---
title: "An Introduction to `mpitbR`"
author: 
  - Girela, Ignacio 
bibliography: references.bib
link-citations: true
date: VIGNETTE UNDER CONSTRUCTION
output: 
  rmarkdown::pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{An introduction to mpitbR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`mpitbR` is a package for calculating Alkire-Foster class measures of multidimensional poverty [@alkire2011; @alkire2015a]. The Alkire-Foster (AF) method distinguishes itself for its versatility in adjusting the indicators, weighting schemes, and poverty cut-offs to different contexts. Indeed, this method is the formal scaffold of the global Multidimensional Poverty Index (MPI) [@alkire2014], a comparably international measure of acute poverty, yearly published by the Oxford Poverty and Human Development Initiative (OPHI) and the United Nations Development Programme (UNDP). In addition, other regional or national MPIs have been created by adapting the global MPI forms to better address local realities.

The global MPI is presented for more than 100 countries, together with ten constituent indicators aligned with Sustainable Development Goals (SDGs), as well as with recommendations of the World Bank’s *Atkinson Commission on Monitoring Global Poverty*. Committed to transparency and collaboration, OPHI publishes all technical files to reproduce their findings. This includes all the Stata do-files to prepare the microdata for generating the global MPI indicators. Then, the global MPI estimates are calculated using the 'mpitb' Stata package developed by Nicolai Suppa [@suppa2023].

The `mpitbR` package faithfully replicates the estimation procedures of the original Stata 'mpitb' package, ensuring methodological consistency for researchers using different programming languages. By offering an R implementation, `mpitbR` contributes to a more integrated and collaborative research ecosystem around the global MPI, aligning with OPHI's encouragement of international collaboration.

This vignette describes basic usage guide for this package, illustrated with real world examples. First, we begin with an introduction to the Alkire-Foster method and the global MPI. subsequently, we demonstrate how to install and start using the `mpitbR` package. For those already familiar to the AF method, can directly proceed to Section 3 and explore multidimensional poverty analysis in practice.

# Multidimensional Poverty Measurement

## The Alkire-Foster method step-by-step

Due to the widely acknowledgement of the multidimensional nature poverty both in academic and policy circles, this century has witnessed an significant emergence of multidimensional poverty measurement methodologies. Among these, the 'dual cut-off' framework proposed by @alkire2011 has gained prominent attention for its flexibility and key-policy properties.

The Alkire-Foster (AF) method can be summarized in the following steps [for a detailed explanation, see @alkire2015a]:

1.  **Establish the data source**

    One of the most salient features of the AF measure is the ability to consider the multiple deprivations faced by the poor jointly. Therefore, all the information ought to come from the same data source, commonly household surveys.

    When designing a multidimensional poverty measure, stakeholders decide which the data source will best align with the poverty measure. As we will see, this selection is linked to two following steps.

2.  **Determine the unit of analysis**

    Depending on the purpose of the MPI in question, the unit of analysis will be defined, i.e., who or what is being studied (individuals, households or even communities). This step influences the choice of indicators, the data source, and interpretation of results. For simplicity, we will refer to 'person' as the unit of analysis.

3.  **Select the dimensions and indicators**

    Poverty is a complex phenomenon, however, for measuring purposes it is necessary to define which dimensions of human development a measure will focus on. Each dimension will be represented by a set of $d \in \mathbb{N}$ indicators (e.g., years of schooling and children school attendance are the two indicators that represents education dimension in the global MPI).

    To represent people's well-being in all dimensions, an $n \times d$ dimensional achievement matrix $X$ is defined, where each element $x_{ij} \in \mathbb{R}_+$ is an ordinal variable that denote the achievement or well-being status of the person $i$ in the $j$-th indicator, for $i = 1,\ldots,n$.

4.  **Define each indicator deprivation cut-off**

    A first cut-off $z_j$ is defined as the minimum level of achievement necessary for being non-deprived in indicator $j$, i.e., if $x_{ij} < z_j$, person $i$ is considered deprived in indicator $j$. We denote the deprivation cut-offs as $z = (z_1, \ldots,z_d)$.

5.  **Obtain the deprivation matrix**

    Then apply the deprivation cut-off vector to each of the $n$ observations to obtain the deprivation matrix $g^0$, where each element is a binary variable denoting the deprivation status of person $i$ in indicator $j$: $g^0_{ij} = 1$ if $x_{ij} < z_j$, and $g^0_{ij} = 0$ if $x_{ij} \geq z_j$. In matrix form,

    $$ g^0 = 
    \begin{bmatrix}
    1 & 0 & 1 & 0 \\
    0 & 1 & 0 & 0 \\
    0 & 1 & 1 & 1 \\
    1 & 0 & 0 & 0 \\
    1 & 0 & 1 & 0 \\
    \end{bmatrix}
    $$ represents the deprivation matrix for five observations and four indicators. The deprivation vector of the first person $g_{1 \cdot } = [1,0,1,0]$ reflects she is deprived in the first and third indicator.

6.  **Assign weights to each indicator**

    The weight $w_j$ of each dimension and indicator reflects their relative importance, where $\sum_{j=1}^d w_j = 1$. In practice, an equal nested weighting scheme is used: dimensions are weighted equally as well as each indicator within the dimension. The weights vector is denoted as $w = (w_1,\ldots,w_d)$.

7.  **Calculate the deprivation score**

    By combining $w$ and $g^0$ is possible to build the weighted deprivation matrix $\bar{g}^0$, where the non-null values of the deprivation matrix $g_{ij}^0$ are replaced by the corresponding value of $w_j$. In matrix form, this is equivalent to $\bar{g}^0 = diag(w) \times g^0$. In the previous example:

    $$ 
    \bar{g}^0 = 
    \begin{bmatrix}
    w_1 & 0 & w_3 & 0 \\
    0 & w_2 & 0 & 0 \\
    0 & w_2 & w_3 & w_4 \\
    w_1 & 0 & 0 & 0 \\
    w_1 & 0 & w_3 & 0 \\
    \end{bmatrix}
    $$

    Then, this information is aggregated by the weighted deprivations to obtain the deprivation score $c_i$ for each person, i.e., $c_i = \sum_{j=1}^{d} w_j g_{i j}^0 = \sum_{j=1}^d \bar{g}_{ij}^0$.

    Let assume that indicators are equally weighted, $w = (1/4,1/4,1/4,1/4)$. Then, the deprivation score vector is $c = (1/2,\, 1/4,\,3/4,\,1/4,\,1/2)$.

8.  **Select the poverty cut-off to identify the poor**

    A second cut-off $k$ is compared with the deprivation score to determine whether the person is poor or not, i.e., if $c_i \geq k$, the $i$-th person is poor. Then, the poverty cut-off $k$ represents the minimum proportion of weighted indicators a person needs to experience to be considered multidimensional poor. This procedure consists of censoring the non-poor from the analysis.

    This identification criterion allows obtaining both the censored (weighted) deprivation matrix, $g^0(k)$, and the censored deprivation score, $c_i(k)$. The former consists on censoring the non-poor from the (weighted) deprivation matrix), i.e., replace the row entries of the non-poor by a vector of zeros. Analogously, the censored deprivation score for each person is equal to $c_i$ if the person is identified as poor, and $0$ otherwise.

    In our example, assume that $k = 1/2$ (i.e., a person is considered poor if she experience deprivation in half or more of the weighted indicators). Then, the second person is not poor and the $g^0(k)$ matrix is expressed as

    $$
    g^0(k) =
    \begin{bmatrix}
    1 & 0 & 1 & 0 \\
    0 & 0 & 0 & 0 \\
    0 & 1 & 1 & 1 \\
    0 & 0 & 0 & 0 \\
    1 & 0 & 1 & 0 \\
    \end{bmatrix}
    $$, and the censored deprivation vector is $c=(1/2,\, 0,\,3/4,\,0,\,1/2)$.

9.  **Calculate the Multidimensional Poverty Index**

    The Multidimensional Poverty Index (MPI) is also referred as the *Adjusted Headcount Ratio*, and denoted as $M_0$. It is defined as the expected value of the censored deprivation score, i.e.,

    $$
    M_0 = \frac{1}{n} \sum_{i=1}^{n} c_i(k)
    $$

    Following the example, $M_0 = \frac{1}{5} (1/2 + 0+3/4+0+1/2) = 0.350$.

## Decomposition properties

The MPI satisfies several desirable properties for a poverty measure. Notably, it allows for valuable decomposition analyses used in policy design. Firstly, the MPI can be disaggregated by two partial measures: incidence and intensity of poverty. These measures enhances the interpretation and understanding of the overall MPI value. Secondly, the MPI can be decomposed by individual indicators. This analysis pinpoints the specific dimensions where poverty is most prevalent, informing policymakers on the most critical areas for intervention. Thirdly, the MPI can be decomposed by population subgroups, such as gender, ethnicity, or rural/urban location. This granularity reveals disparities in poverty experiences across different segments of society, guiding targeted interventions.

### Incidence and intensity of poverty

The $M_0$ measure can be expressed as the product of two partial indices representing the incidence and intensity of poverty. Recall that people are identified as poor if $c_i \leq k$. Let $q$ denote the number of people identified as poor. Then, we can multiply and divide $M_0$ by $q$ and rearrange some terms to compute the incidence and intensity.

$$
M_0 = \frac{1}{n} \sum_{i=1}^{n} c_i(k) \times \frac{q}{q} = \frac{q}{n} \times \frac{1}{q} \sum_{i=1}^{n} c_i(k) = H \times A
$$ , where $H$ and $A$ denote the incidence and intensity of poverty, respectively.

The incidence of poverty $H$ represents the proportion of multidimensional poor people in a society and it is represented in percentage terms. Recapitulate our previous example. In that case, the first, third, and the fifth person were identified as poor, $q$, over all the three people $n=5$. Then, the incidence is equal to $H=3/5=0.6$, i.e., $60.0\%$ of the population is multidimensional poor.

On the other hand, the intensity of poverty $A$ represents the average weighted deprivations (score) that the poor experience. In our example, $A = (1/2+ 0+3/4+0+1/2)/3 = 58.33\%$, i.e., on average, the poor people is deprived in $58.33\%$ of the weighted indicators.

Here is why the MPI is also named as *Adjusted Headcount Ratio*: it represents the proportion of multidimensionally poor people, adjusted by the average intensity they experience. Notably, if all the poor face deprivations across all indicators, the MPI would be equal to the incidence $H$ of poverty. Then, if we multiply $H$ by the intensity $A$, $M_0$ now represents the proportion of weighted deprivations that the poor experience within a society out of the total potential deprivations they could experience overall.

### Indicators breakdown

A relevant property of the MPI for policy analysis is its decomposability by censored indicators. Note that in the step-by-step procedure, we construct the MPI by first aggregating information across censored indicators by columns and then across individuals by rows. The same result can be achieved by reversing this aggregation order.

To see this, take the $M_0$ equation and rearrange the aggregation order, i.e.,

$$
M_0 = \frac{1}{n} \sum_{i=1}^{n} c_i(k) =\frac{1}{n} \sum_{i=1}^{n} \left[ \sum_{j=1}^d w_j g_{ij}^0(k) \right] = \sum_{j=1}^{d} w_j \left[\frac{1}{n} \sum_{i=1}^n g_{ij}^0(k) \right]
$$ where, we define $h_j(k) = \frac{1}{n} \sum_{i=1}^n g_{ij}^0(k)$ as the censored headcount ratio of indicator $j$. Similarly, the uncensored headcount ratio is calculated using the uncensored deprivation matrix: $h_j= \frac{1}{n} \sum_{i=1}^n g_{ij}^0$. The MPI can be expressed as the sum of the censored headcount ratio of all indicators, each weighted by their corresponding relative weight, $w_j$, i.e.,

$$
M_0 = \sum_{j=1}^{d} w_j\, h_j(k)
$$ In our example, the uncensored indicator headcounts are $h = (60\%,\, 40\%,\,60\%,\,20\%)$, while the censored indicator headcounts are $h(k) = (40\%,\,20\%,\,60\%,\,20\%)$. Comparing the difference between the uncensored and censored indicators yields valuable insights into the prevalence of deprivation among the poor population.

On the other hand, the absolute and percentage contribution of each indicator is reported. The absolute contribution of each indicator is determined by multiplying its weight by the censored indicator value. The percentage contribution is calculated as follows:

$$\phi_j = w_j \frac{h_j(k)}{M_0}$$ where $\phi_j$ is the relative contribution of indicator $j$ in the Adjusted Headcount Ratio. Following the example, the percentage contributions are $28.58\%, 14.29\%, 42.85\%, 14.29\%$, which sum to 1.

### Subgroup Decomposition

Another key policy-relevant property of the MPI is the subgroup decomposition. The $M_0$, and other associated AF measures ($H$, $A$, $h_j$, $h_j(k)$), can be calculated for various population subgroups, such as age cohorts, regions, living areas, gender, ethnicity, and educational attainment. This allows the overall measure to be expressed as the sum of the measures by group, weighted by the share of the population of that subgroup. Formally,

$$
M_0 = \sum_{l=1}^{L} \nu_l\, M_0^{l}
$$ where $\nu_l = n_l / n$ is the proportion of people belonging to the population subgroup $l$, and $M_0^l$ is the Adjusted Headcount Ratio of the population subgroup $l$.

In our example, consider a population divided into two regions: North and South, where first two individuals are from the North region, while the remaining individuals are from the South region. The population share of the South region is greater than the North ($60\%$ vs. $40\%$). The MPI for the North region is $0.250$, whereas the South region exhibits a higher MPI of $0.416$.

## Changes over time

Over the past decade, there has been a significant increase in the frequency of multi-topic surveys, enabling the calculation of multidimensional poverty measures. @COT-mpi introduced four measures to evaluate pro-poor multidimensional poverty reduction between two periods of time using repeated cross-sectional data. These measures are initially explained using the Adjusted Headcount Ratio $M_0$, but their application can be extended to the other AF measures: $H$, $A$, $h_j$, and $h_j(k)$.

Let $t^1$ and $t^2$ denote the initial and final periods, respectively. $M_0^{t^1}$ and $M_0^{t^2}$ represent their corresponding Adjusted Headcounts Ratio. Importantly, for comparative purposes, these two poverty measures must be calculated using the same set of parameters: identical indicators, weights, deprivations thresholds and poverty cutoffs.

To quantify changes in the MPI, we employ the following measures:

Firstly, we utilize the *absolute rate of change* ($\Delta$), calculated as the difference in $M_0$ between the final period $t^2$ and the initial period $t^1$:

$$
\Delta(M_0) = M_0^{t^2} - M_0^{t^1} 
$$

Secondly, we emply the *relative rate of change* ($\delta$), which expresses the change in $M_0$ as a percentage of the initial level $t^1$:

$$
\delta(M_0) = \frac{M_0^{t^2} - M_0^{t^1}}{M_0 ^{t^1}} \times 100
$$

To facilitate comparisons of changes in M0 over time across countries with varying reference periods, annualized versions of the previously introduced measures are employed.

The *annualized absolute rate of change* ($\bar{\Delta}$) is derived by dividing the absolute rate of change ($\Delta$) by the difference in time periods ($t^2 - t^1$):

$$
\bar{\Delta}(M_0) = \frac{M_0^{t^2} - M_0^{t^1}}{(t^2 - t^1)}
$$

Finally, the *annualized relative rate of change* ($\bar{\delta}$) represents the compound rate of change in $M_0$ per year between the initial and final periods. It is calculated as:

$$
\bar{\delta}(M_0) = \left[ \left( \frac{M_0^{t^2}}{M_0^{t^1}} \right)^{\frac{1}{t^2 - t^1}} - 1 \right]
$$

These annualized measures provide a standardized framework for assessing the pace of change in $M_0$ across different contexts, enabling more meaningful comparisons and analyses.

## The global MPI

### Components

We have previously introduced the global MPI, an international measure of acute multidimensional poverty, aligned with the Sustainable Development Goals (SDGs). It comprises ten deprivation indicators grouped into three poverty dimensions: two for health, two for education, and six for living standards. Their weights and deprivation cut-offs are defined as follows:

-   **Health** (1/3 weight)

    -   [Child Mortality]{.underline} (1/6 weight): Deprived if any child in the household has died in the five years preceding the survey.

    -   [Nutrition]{.underline} (1/6 weight): Deprived if any child under five years old is underweight or any adult is undernourished.

-   **Education** (1/3 weight)

    -   [Years of Schooling]{.underline} (1/6 weight): Deprived if no household member aged 10 years or older has completed at least six years of schooling.

    -   [School Attendance]{.underline} (1/6 weight): Deprived if any school-aged child is not currently enrolled in school.

-   **Living Standards** (1/3 weight)

    -   [Cooking Fuel]{.underline} (1/18 weight): Deprived if the household relies on solid fuels (such as wood, dung, or charcoal) for cooking.

    -   [Sanitation]{.underline} (1/18 weight): Deprived if the household lacks access to improved sanitation facilities.

    -   [Drinking Water]{.underline} (1/18 weight): Deprived if the household lacks access to improved drinking water sources or if improved drinking water is more than a 30-minute round trip to collect.

    -   [Electricity]{.underline} (1/18 weight): Deprived if the household lacks access to electricity.

    -   [Housing]{.underline} (1/18 weight): Deprived if the household's housing structure is inadequate (e.g., natural materials for walls, floors, or roofs).

    -   [Assets]{.underline} (1/18 weight): Deprived if the household does not own more than one of the following assets: radio, television, telephone, computer, animal cart, bicycle, motorbike, or refrigerator.

The weighting methodology employs nested equal weights. All dimensions are assigned equal weights, and within each dimension, all indicators are considered equally relevant. In the global MPI, each of the three dimensions (Health, Education, and Living Standards) receives a weight of one third.

Within the Health and Education dimensions, each indicator receives an equal weight of 1/2. Since the dimension carries a weight of 1/3, each indicator within that dimension has an overall weight of 1/6.

Analogously, each indicator of the Living Standards dimension receives a 1/6 weight, resulting in an overall weight of 1/18 for each indicator within that dimension.

The global MPI establishes a poverty cut-off of $k =1/3$ to measure acute poverty. A person is considered multidimensionally poor if they are deprived in $33.33\%$ or more of the weighted indicators. Alternatively, given the weight assigned to each dimension, experiencing deprivation in at least one dimension can also be interpreted as being poor. Furthermore, the global MPI reports results using poverty cut-offs of $20\%$ and $50\%$ to denote vulnerability and severe poverty status, respectively.

Finally, the annual global MPI report includes other key information such as the indicator breakdown and percentage contributions (including uncensored levels of deprivation in each indicator), disaggregated details by certain population subgroups (rural-urban, age groups, and subnational regions), and other key estimates for inference (standard errors and confidence intervals).

### Unit of identification and unit of analysis

When employing the MPI, a critical distinction lies between the unit of identification and the unit of analysis. These concepts significantly influence how poverty is measured and understood within the MPI framework, with substantial implications for poverty analysis

The unit of identification is the entity from which data is collected for poverty assessment. This could be an individual, a household, or even a community. The choice of unit carries crucial assumptions. For instance, selecting the household as the unit implies that poverty affects all members of the household equally. Since it uses available information on all household members, the global MPI utilizes the household as the unit of identification.

The unit of analysis refers to the level at which data is aggregated and analyzed to draw conclusions about poverty. While the household might be the unit of identification, the global MPI analyzes data at the individual level using appropriate sampling weights. Reporting results by individuals facilitates the exploration of gendered or age-related disparities and enables the examination of intra-household variations in poverty experience.

In conclusion, the careful consideration of both the unit of identification and the unit of analysis is paramount for robust and insightful poverty assessments using the MPI.

### Main data sources

The MPI uses information from three main sources that are publicly available and consistent for most developing countries. These sources are:

-   The Demographic and Health Surveys ([DHS](https://dhsprogram.com/))
-   The Multiple Indicators Cluster Survey ([MICS](https://mics.unicef.org/))
-   National Household Surveys: If information from the former surveys is not available for a specific country, the MPI may use data from other surveys conducted within that country, as long as those surveys cover the same topics. For instance, data from the national survey 'Encuesta de Condiciones de Vida 2013-2014' was used to calculate the global MPI in Ecuador.

Every year, the global MPI report, country briefings, data tables and technical files are updated. In some cases, indicators definitions are refined. All particular national variations are documented in the methodological notes for the year in which the MPI was released. The major revision was in 2018 to align the indicators with SDGs.

# The `mpitbR` package

Having established the fundamental concepts in the previous section, we now delve into the core of this vignette, focusing on the measurement of multidimensional poverty. We will do so while carefully considering best practices and relevant considerations.

## Installation

The simplest way to install `mpitbR` is to download and install it directly from CRAN by typing the following command in R console:

```{r, eval=FALSE}
install.packages("mpitbR")
```

Another way is to install the development version from the `mpitbR` Github repository:

```{r, eval=FALSE}
library(devtools)

install_github("girelaignacio/mpitbR")
```

## Data processing

This vignette utilizes two preprocessed Benin DHS surveys (DHS06 and DHS17-18) included in this package. These datasets were prepared using Stata do-files to process the raw microdata according to the specifications outlined by the OPHI and the UNDP for their annual global MPI reports.

These datasets include two unique identifiers (household and individual IDs), variables related to the survey design (primary sampling unit, stratum, and sampling weight), demographic characteristics (sex, living area, and region), and ten columns representing each of the ten indicators used in the global MPI calculation. The columns `d_cm, d_nutr, d_satt, d_educ, d_elct, d_wtr, d_sani, d_hsg, d_ckfl, d_asst` represents Child Mortality, Nutrition, School Attendance, Years of Schooling, Electricity, Water, Sanitation, Housing, Cooking Fuel, and Assets, respectively.

### Load package and data

Below, we load the installed `mpitbR` package and present a few rows of the preprocessed Benin 2006 round DHS microdata, \`ben_dhs06\`:

```{r}
library(mpitbR)

head(ben_dhs06)
```

**!** Note that the `mpitbR` package operates directly on the uncensored deprivation matrix, where all the indicators columns contain binary values (0 and 1). Consequently, the package cannot be used for earlier stages of the analysis, such as generating deprivation indicators or making normative decisions regarding their definition.

### Missing values

Multidimensional poverty indicators often contain missing values due to non-response from household members. Missing data are more common in cohort-specific deprivations, where the unit of analysis is the individual. For example, when the unit of analysis is the household, if a child's anthropometric measurements are missing, not only is the child assigned a missing value, but it also impacts the poverty status of the entire household.

To ensure accurate MPI calculations, an important step involves verifying that all missing values within indicators are consistently assigned to all members of the unit of identification (e.g., the household). On the other hand, data cleaning should focus on retaining only the relevant columns, including those containing survey design data (e.g., primary sampling units (PSUs), weights, strata), and variables related to population groups, as it in `ben_dhs06` data.

The code below explore the total number of missing values for all the indicators columns in `ben_dhs06` data using `tidyverse` package.

```{r tidyverse, message=FALSE}
# Load `tidyverse` package
library(tidyverse)

# Count missing values by all the deprivation indicators columns 
  # (all their names start with d_*)
indicators_NAs <- ben_dhs06 %>% 
  summarise(across(grep("^d_",colnames(ben_dhs06)),  ~sum(is.na(.))))

print(indicators_NAs)

# Now compare the total number of missing values in the dataset with the indicators. 
total_NAs <- sum(is.na(ben_dhs06))
print(total_NAs == sum(indicators_NAs))
```

The `d_nutr` and `d_cm` columns represent deprivation indicators for nutrition and child mortality, respectively. Within the education dimension, `d_satt` and `d_educ` correspond to deprivation indicators for children's school attendance and years of schooling, respectively. Finally, the following columns represent deprivation indicators for: electricity access (`d_elct`), improved sanitation (`d_sani`), water access (`d_wtr`), housing quality (`d_hsg`), cooking fuel (`d_ckfl`), and asset ownership (`d_asst`).

We observe a higher frequency of missing values within the Health dimension indicators compared to other dimensions. Given that these missing values are exclusively associated with deprivation indicators, we can employ the `na.omit()` function to directly remove observations containing any missing data.

```{r na.omit}
ben_dhs06 <- na.omit(ben_dhs06)
```

**!** If one observation (unit of analysis) has a missing value, all other observations belonging to the same group (unit of identification) exhibit a missing value for that variable. While OPHI do-files prevent this to occur, practitioners should remain attentive to such inconsistencies in their measurement projects.

### Household survey design

Household surveys, the primary data source for multidimensional poverty measurement, employ complex survey designs. To ensure the reliability of point estimates and their associated standard errors and confidence intervals -crucial for statistical inference- `mpitbR` accounts for complex survey designs by utilizing methods from the `survey` R package [@survey] to account for these designs. This is another reason why it is important to remove missing values, as they can introduce subtle and potentially difficult-to-detect biases into the estimates generated by `survey` package functions.

We know define the survey design using the `svydesign` function from the `survey` package, considering the primary sampling units (`psu`), sampling weights (`weight`), and strata (`strata`) information in the data

```{r survey, message=FALSE}
# Load `survey` library
library(survey)

# Define the survey design
svydata <- svydesign(ids = ~psu, weights = ~weight, strata = ~strata, data = ben_dhs06)
```

## Define the multidimensional poverty measurement project

Once the survey design is set, we specify the MPI measurement project settings. This includes defining our data source (`svydata` object in this case), identifying the dimensions and assigning indicator columns to each dimension, and optionally providing a label for our project (a brief name with a short description). To accomplish this, we utilize the `mpitb.set` function.

Since we are reproducing the global MPI for Benin DHS06, we group our indicator columns in health, education and living standards dimensions using a list.

```{r indicators}
# Group indicators by dimension
indicators <-  list(hl = c("d_nutr","d_cm"),
                    ed = c("d_satt","d_educ"),
                    ls = c("d_elct","d_sani","d_wtr","d_hsg","d_ckfl","d_asst"))
```

Next, pass the data and indicators as arguments to the `mpitb.set` function.

```{r set_proj}
# Set the multidimensional poverty project
set <- mpitb.set(data = svydata, indicators = indicators,
                 name = "ben_dhs06", desc = "Benin global MPI 2006")
```

`set` is an object of class `mpitb_set` that contains all the relevant information of the MPI measurement project for further use.

## Cross-sectional estimates

The core of this package is `mpitb.est`, designed for estimating the MPI and their partial measures (intensity, incidence, indicator-specific measures by population subgroups).

The `mpitb.est` function offers several arguments to allow for customized MPI calculations. Here below, we outline the key arguments (for a comprehensive list of arguments and their descriptions, type `?mpitb.est` in your R console):

- `set` is the multidimensional poverty measurement project settings previously defined with `mpitb.set`.  

- `k` is the vector of poverty cut-offs (values between 1 and 100). 

- `weights` is for vector specifying the weighting scheme for each dimension. By default, equal nested weights are used, as employed in the global MPI.   

- `measures` refers to the main aggregate measures to be calculated ($M_0$, $H$, or $A$). 

- `indmeasures` are all the indicator-specific measures, such as censored and raw headcount ratios, and their absolute and percentage contribution to overall poverty.

By default, all measures are estimated. 

### A quick start

As an example, we will now demonstrate the estimation of the global MPI for Benin in 2006.

```{r first_globalMPI}
# Estimate the Benin global MPI 2006
estimate.01 <- mpitb.est(set, k = 33, measures = "M0", indmeasures = NULL)
```
Upon execution, the `mpitb.est` function displays a message that includes the function call itself, a list of dimensions with their assigned indicators and corresponding weights (allowing users to verify the setup), the specific measures that have been estimated, relevant estimation parameters such as the number of poverty cut-offs and subgroups analyzed, and other important features like the confidence level used for calculating confidence intervals and whether parallel estimation has been employed. This detailed message can be suppressed by setting the `verbose` argument to `FALSE` within the `mpitb.est` function call.

The estimation results are stored in the `estimate.01` object, which is an instance of the `mpitb_est` class. This object is a list containing two data frames: `lframe`, which encompasses all cross-sectional estimates for each level of analysis, and `cotframe`, which contains measures of change over time for each level of analysis. However, since this example does not involve time-series analysis, the `cotframe` element will be `NULL`. This structure allows for flexible storage and retrieval of both cross-sectional and longitudinal poverty estimates.

```{r head_first_globalMPI}
# Take a glance at the results
as.data.frame(estimate.01$lframe)
```

This displays the raw data frame of our estimates. The first four columns provide the point estimate (`b`), the standard error of the point estimate (`se`), which accounts for the survey design, and the lower and upper bound of the confidence interval (`lb` and `ub`), with confidence intervals calculated considering the measures as proportions. `measure` and `k` columns indicate the specific AF measure and poverty cut-off, respectively, to which each estimate corresponds.

### Analyzing poverty across population groups

We have previously established that the MPI can be calculated as the sum of the MPIs of mutually exclusive population subgroups, weighted by the population share of each group. This decomposition is crucial for identifying sociodemographic disparities in poverty distribution. Typically, global MPI analyses explore differences across rural-urban areas, different age cohorts, and by gender.

To calculate the MPI for specific subgroups, utilize the `over` argument within the `mpitb.est` function. This argument accepts a character vector specifying the column names of the population groups for which poverty analysis is desired.

To illustrate, let's examine disparities between rural and urban areas in Benin during 2006. We can execute the following code:
 
```{r rural-urban}
# Include living areas in the MPI calculation
estimate.02 <- mpitb.est(set, k = 33, measures = "M0", indmeasures = NULL, 
                         over = "area", verbose = FALSE)

# View results
as.data.frame(estimate.02$lframe)
```

Again, results are presented as a data frame. The population group (in this case, `area`) is identified in `loa` column, which stands for 'level of analysis'. The corresponding MPI estimates for each level of analysis are labeled by each subgroup in the `subg` column. Note that another `loa` entry exists with a unique subgroup called "nat". This represents the MPI estimate calculated across all the observations within the data set, generally at the national level. To exclude the overall national estimate, set the `overall` argument to `FALSE` within the `mpitb.est` function.

The following code generates a barplot using `ggplot2` to visualize and compare national, and living-area poverty levels.

```{r plot_rural-urban, fig.cap="Multidimensional Poverty by Living Areas in Benin 2006",fig.width=3, fig.height=3, fig.align="center"}
ggplot(as.data.frame(estimate.02$lframe), aes(x = subg, y = b, fill = subg)) +
    geom_bar(stat = "identity", position = "dodge", width = 0.5) +
    geom_errorbar(aes(ymin = ll, ymax = ul), width = 0.1) +
    labs(x = "Level of Analysis", y = "MPI", fill = "Subgroups") +
    theme_bw() +
    theme(legend.position = "bottom") 
```

Figure \@ref(fig:plot_rural-urban) demonstrates that poverty levels in rural areas are significantly higher than in urban areas of Benin ($0.531$ and $0.285$, respectively). This suggests that, on average, individuals experiencing poverty in rural regions exhibit a higher degree of multidimensional poverty compared to their urban counterparts.

Confidence intervals are displayed at the top of each bar. A good practice for comparing poverty levels is to examine if the confidence intervals do not overlap. In this example, it is possible to infer that multidimensional poverty in rural Benin is statistically higher than in urban Benin. 

To further investigate whether higher multidimensional poverty in rural Benin is primarily driven by a larger proportion of poor individuals (incidence, $H$) or by the rural poor experiencing greater deprivation (intensity, $A$), we can incorporate these two measures into our previous analysis using the `measures` argument.

```{r rural-urban2}
# Include incidence H and intesity A in the MPI calculation
estimate.03 <- mpitb.est(set, k = 33, measures = c("M0","H","A"), indmeasures = NULL, 
                         over = c("area"), verbose = FALSE)

# Explore coefficients of H and A
  # Incidence
coef(subset(estimate.03$lframe, measure == "H" & loa == "area"))
 # Intensity
coef(subset(estimate.03$lframe, measure == "A" & loa == "area"))
```
While the intensity of poverty is statistically higher in rural areas of Benin than in urban areas, the difference in intensity compared to urban areas is not as pronounced as the difference in incidence ($87.59\%$ of the rural population experiencing multidimensional poverty compared to $52.73\%$ in urban areas).

**!** A final important caveat: when analyzing multidimensional poverty across different subgroups, avoid subsetting the dataset to isolate specific groups before specifying the measurement project with `mpitb.set` function. Subsetting the data can impact the degrees of freedom, potentially compromising the accuracy of statistical inferences. 

### Indicator-specific measures analysis

Breaking down the MPI by indicators provides valuable insights in multidimensional poverty analysis. We can compare the censored and uncensored headcount ratios of each indicator and explore their contribution (absolute and percentage) to the MPI. This granular analysis can be further refined by examining these contributions across different population subgroups, offering a high-resolution lens of poverty within a society.

As mentioned earlier, the argument `indmeasures` in the `mpitb.est` function encompasses all the indicator-specific measures and, by default, all of them are calculated. This is why we previously set this argument to `NULL`. 

```{r indicators_measures}
# Estimate indicator-specific measures
  # We specify nothing in `indmeasures`
  # since all of them are calculated by default. 
estimate.04 <- mpitb.est(set, k = 33, measures = "M0", 
                         over = "area")

# View results
head(estimate.04$lframe)
```


## Changes over time

# Appendix

## Appendix 1: Moving from Stata to R

This appendix will contain some comparisons between estimates in the R and Stata version of the package.

The equivalent command of the previous example in the 'mpitb' Stata package is the following:

``` stata
mpitb est, name(trial01) measures(all) indmeasures(all) aux(all) \\\ 

  klist(20 33) weight(equal) lframe(myresults, replace) \\\ 

  svy over(area region) cotmeasures(all) cotframe(mycot, replace) tvar(t) cotyear(year)
```

Here below, we compare the results from the Stata package and this version in R. We merged both data frames with the results (point estimate, standard errors, and confidence intervals). We calculate the L1 distance between the estimates ($|\theta_{R}-\theta_{Stata}|$) by measure in the case of the "lframe" results and by ctype and annualized measure in the case of the "cotframe" results.

```{r read_the_dataframes, echo=FALSE, eval=FALSE}

# lframe testing ----------------------------------------------------------

R <- est$lframe
R <- R[,-which(colnames(R) == "ctype")]

Stata <- as.data.frame(haven::read_dta("~/lframe.dta"))

Stata <- Stata[, !(names(Stata) %in% c("spec","wgts","pval","tval"))]
Stata$subg[which(is.na(Stata$subg))] <- "nat"
Stata <- Stata[-( which(Stata$measure == "popsh")),]
Stata$indicator[which(Stata$indicator== "")] <- NA
Stata <- Stata[,-which(colnames(Stata) == "ctype")]
Stata <- Stata[names(R)]

R<-R[order(R$subg, R$t, R$loa,R$measure,R$k,R$indicator),]
Stata<-Stata[order(Stata$subg, Stata$t, Stata$loa,Stata$measure,Stata$k,Stata$indicator),]

merged <- merge(R,Stata,by= c("subg","t","loa","measure","k","indicator"))



# cotframe testing --------------------------------------------------------


cotR <- est$cotframe
cotR <- cotR[,-(which(colnames(cotR) == "yt0")) ]
cotR <- cotR[,-(which(colnames(cotR) == "yt1")) ]

cotStata <- as.data.frame(haven::read_dta("~/cotframe.dta"))
cotStata <- cotStata[, !(names(cotStata) %in% c("spec","wgts","pval","tval","yt0","yt1","t0","t1"))]
cotStata$subg[which(is.na(cotStata$subg))] <- "nat"
cotStata$indicator[which(cotStata$indicator== "")] <- NA
cotStata$ctype <- haven::zap_labels(cotStata$ctype)
cotStata$ctype[which(cotStata$ctype == 1)] <- "abs"
cotStata$ctype[which(cotStata$ctype == 2)] <- "rel"


cotStata <- cotStata[names(cotR)]

cotR<-cotR[order(cotR$subg, cotR$loa,cotR$measure,cotR$k,cotR$indicator,cotR$ctype,cotR$ann),]
cotStata<-cotStata[order(cotStata$subg, cotStata$loa,cotStata$measure,cotStata$k,cotStata$indicator, cotStata$ctype, cotStata$ann),]


cot.merged <- merge(cotR,cotStata,by= c("subg","loa","measure","k","indicator","ctype","ann"))

```

The following output shows the summary of the distribution of this comparisons (suffix ".x" refers to the estimates of the R package and ".y" to the Stata version).

```{r summaries, echo=T, eval=FALSE}
# "lframe" COMPARISONS
aggregate(abs(b.x - b.y)~ measure, merged, summary)
aggregate(abs(se.x - se.y)~ measure, merged, summary)
aggregate(abs(ll.x - ll.y)~ measure, merged, summary)
aggregate(abs(ul.x - ul.y)~ measure, merged, summary)
# "cotframe" COMPARISONS
aggregate(abs(b.x - b.y)~ ctype+ann, cot.merged, summary)
aggregate(abs(se.x - se.y)~ ctype+ann, cot.merged, summary)
aggregate(abs(ll.x - ll.y)~ ctype+ann, cot.merged, summary)
aggregate(abs(ul.x - ul.y)~ ctype+ann, cot.merged, summary)

```

All the results are very close with some exceptions found in few estimates of the standard errors of the relative changes over time measures. Since the point estimates ("b") are close and Delta method is used for calculating the standard errors, here there may be some differences between the "survey" library in R and Stata.

## Appendix 2: Considerations on convergence in some estimates

## Appendix 3: Speed-up procedures with parallel computing

As previously noted, the total number of estimates can rapidly increase as we consider more subgroups, poverty lines, and other factors. However, most of these estimates are independent of each other. This characteristic allows for the parallelization of many computations, significantly improving processing speed.

While parallel computing has traditionally been more prevalent in server environments, the widespread adoption of multi-core processors in modern personal computers has significantly enhanced the accessibility of parallel computing for users at all levels, enabling more efficient computation.

# References
