---
title: "Moving from Stata to R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{StataComparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mpitbR)
```


## Some comparisons with the original Stata package

The equivalent command of the previous example in the <tt>mpitb</tt>
Stata package is the following:

> mpitb est, name(trial01) measures(all) /// indmeasures(all) aux(all)
> /// klist(20 33) weight(equal) /// lframe(myresults, replace) /// svy
> over(area region) /// cotmeasures(all) cotframe(mycot, replace)
> tvar(t) cotyear(year)

Here below, we compare the results from the Stata package and this
version in R. We merged both data frames with the results (point
estimate, standard errors, and confidence intervals). We calculate the
L1 distance between the estimates ($|\theta_{R}-\theta_{Stata}|$) by
measure in the case of the "lframe" results and by ctype and annualized
measure in the case of the "cotframe" results.

```{r read_the_dataframes, echo=FALSE, eval=FALSE}

# lframe testing ----------------------------------------------------------

R <- est$lframe
R <- R[,-which(colnames(R) == "ctype")]

Stata <- as.data.frame(haven::read_dta("~/lframe.dta"))

Stata <- Stata[, !(names(Stata) %in% c("spec","wgts","pval","tval"))]
Stata$subg[which(is.na(Stata$subg))] <- "nat"
Stata <- Stata[-( which(Stata$measure == "popsh")),]
Stata$indicator[which(Stata$indicator== "")] <- NA
Stata <- Stata[,-which(colnames(Stata) == "ctype")]
Stata <- Stata[names(R)]

R<-R[order(R$subg, R$t, R$loa,R$measure,R$k,R$indicator),]
Stata<-Stata[order(Stata$subg, Stata$t, Stata$loa,Stata$measure,Stata$k,Stata$indicator),]

merged <- merge(R,Stata,by= c("subg","t","loa","measure","k","indicator"))



# cotframe testing --------------------------------------------------------


cotR <- est$cotframe
cotR <- cotR[,-(which(colnames(cotR) == "yt0")) ]
cotR <- cotR[,-(which(colnames(cotR) == "yt1")) ]

cotStata <- as.data.frame(haven::read_dta("~/cotframe.dta"))
cotStata <- cotStata[, !(names(cotStata) %in% c("spec","wgts","pval","tval","yt0","yt1","t0","t1"))]
cotStata$subg[which(is.na(cotStata$subg))] <- "nat"
cotStata$indicator[which(cotStata$indicator== "")] <- NA
cotStata$ctype <- haven::zap_labels(cotStata$ctype)
cotStata$ctype[which(cotStata$ctype == 1)] <- "abs"
cotStata$ctype[which(cotStata$ctype == 2)] <- "rel"


cotStata <- cotStata[names(cotR)]

cotR<-cotR[order(cotR$subg, cotR$loa,cotR$measure,cotR$k,cotR$indicator,cotR$ctype,cotR$ann),]
cotStata<-cotStata[order(cotStata$subg, cotStata$loa,cotStata$measure,cotStata$k,cotStata$indicator, cotStata$ctype, cotStata$ann),]


cot.merged <- merge(cotR,cotStata,by= c("subg","loa","measure","k","indicator","ctype","ann"))

```

The following output shows the summary of the distribution of this
comparisons (suffix ".x" refers to the estimates of the R package and
".y" to the Stata version).

```{r summaries, echo=T, eval=FALSE}
# "lframe" COMPARISONS
aggregate(abs(b.x - b.y)~ measure, merged, summary)
aggregate(abs(se.x - se.y)~ measure, merged, summary)
aggregate(abs(ll.x - ll.y)~ measure, merged, summary)
aggregate(abs(ul.x - ul.y)~ measure, merged, summary)
# "cotframe" COMPARISONS
aggregate(abs(b.x - b.y)~ ctype+ann, cot.merged, summary)
aggregate(abs(se.x - se.y)~ ctype+ann, cot.merged, summary)
aggregate(abs(ll.x - ll.y)~ ctype+ann, cot.merged, summary)
aggregate(abs(ul.x - ul.y)~ ctype+ann, cot.merged, summary)

```

All the results are very close with some exceptions found in few
estimates of the standard errors of the relative changes over time
measures. Since the point estimates ("b") are close and Delta method is
used for calculating the standard errors, here there may be some
differences between the "survey" library in R and Stata

------------------------------------------------------------------------
